---
title: "Using Constellation to Identify Sepsis"
author: "Mark Sendak"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identify Sepsis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Sepsis

Implementing a sepsis definition motivated development of the constellation package. There is ongoing debate in the medical literature about what constitutes sepsis, but our local team of clinical experts agreed to define sepsis as the occurrence of all 3 following criteria:

1. 2 or more SIRS criteria  
    + Temperature > 100.4 C or Temperature < 96.8 C (**6 hours**)
    + Heart Rate > 90 (**6 hours**)
    + Respiration Rate > 20 (**6 hours**)
    + WBC Count > 12, WBC Count < 4 (**24 hours**)
2. Blood culture order (**24 hours**)
3. End organ damage  
    + Creatinine > 2.0 (**24 hours**)
    + INR > 1.5 (**24 hours**)
    + Total bilirubin > 2.0 (**24 hours**)
    + systolic BP < 90 or decrease in SBP by > 40 over 6 hours (**6 hours**)
    + Platelets < 100 (**24 hours**)
    + Lactate >= 2 (**24 hours**)

For each criteria, there is both a threshold to signify dysfunction as well as a relevant time window to search for events. A major challenge of working with medical data is that the sampling rate varies across measurements. For example, vital signs are sampled more frequently than blood labs, so their information is carried forward for a shorter period of time than lab results.

There are 3 steps to identify sepsis and each step corresponds to a constellation function:

1. Identify instances of systolic blood pressure drops > 40 over 6 hours (**value_change()** function)
2. Identify instances of SIRS >= 2 (**constellate_criteria()** function)
3. Identify instances in which patients have SIRS >= 2, a blood culture order, and evidence of end organ damage (**constellate()** function)

First, load constellation and fasttime
```{r setup, message = FALSE}
devtools::install_github("marksendak/constellation")
library(constellation)
library(fasttime)
```

## value_change() - identify systolic blood pressure drops

Use this function to identify all instances of a drop in systolic blood pressure of greater than 40 within 6 hours.

```{r bp_drop}
systolic_bp <- vitals[VARIABLE == "SYSTOLIC_BP"]
systolic_bp[, RECORDED_TIME := fastPOSIXct(RECORDED_TIME)]

systolic_bp_drop <- value_change(systolic_bp, value = 40, direction = "down",
    window_hours = 6, join_key = "PAT_ID", time_var = "RECORDED_TIME", 
    value_var = "VALUE", mult = "all")
head(systolic_bp_drop)
```

## constellate_criteria() - identify SIRS >= 2

First, break out all the SIRS component tables applying dysfunction thresholds
```{r sirs_components}
temp <- vitals[VARIABLE == "TEMPERATURE" & (VALUE > 100.4 | VALUE < 96.8)]
pulse <- vitals[VARIABLE == "PULSE" & VALUE > 90]
resp <- vitals[VARIABLE == "RESPIRATORY_RATE" & VALUE > 20]
wbc <- labs[VARIABLE == "WBC" & (VALUE > 12 | VALUE < 4)]
```

Next, identify which criteria are met at every measurement time
```{r sirs_calculate}
sirs <- constellate_criteria(temp, pulse, resp, wbc, criteria_names = c("TEMPERATURE", "PULSE", "RESPIRATORY_RATE", "WBC"), window_hours = c(6, 6, 6, 24), join_key = "PAT_ID", time_var = "RECORDED_TIME")
head(sirs)
```

Sum values in the 4 columns to calculate SIRS score at every measurement time and subset to instances where SIRS >= 2
```{r sirs_sum}
sirs[, SIRS_SCORE := TEMPERATURE + PULSE + RESPIRATORY_RATE + WBC]
sirs <- sirs[SIRS_SCORE >= 2]
head(sirs)
```

## constellate() - identify instances where 3 sepsis criteria are met

#### Compile end organ damage

#### Run constellate()


## Technical References

If you'd like to learn more about our sepsis model, please see our publications from the [International Conerence on Machine Learning](https://arxiv.org/abs/1706.04152) and [Machine Learning in Health Care](https://arxiv.org/pdf/1708.05894.pdf). 
